---
title: "03_PreProcessing_Lecture.Rmd"
author: "Morrow Zhang"
date: "`r Sys.Date()`"
output: html_document
---

# Goals
Goal of this analysis is to remove any ASVS from chloroplasts and mitochondria. Evaluate negative and positive controls while removing the negative control. Check for reverse complements and sequencing depth of samples. Add the metada for my dataset and write a preprocessed_physeq for downstream workflow.

# Set up the Environment 

```{r }

# Set seed
set.seed(238428)

# Time of start
start_time <- Sys.time()

# Load packages 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, patchwork,
               # Note to add ggpubr & rstatix!!
               ggpubr, rstatix, install = FALSE)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.path = "figures/03_PreProcessing/"
)
```


# Load Phyloseq Data 

```{r}
load("data/raw_physeq.RData")

raw_physeq
```
# Clean up the Data 

## Remove Chloroplasts and Mitochondria 

```{r}
noChloros_physeq <- 
  raw_physeq %>% 
  subset_taxa(Order != "Chloroplast" | is.na(Order))

num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs

noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs
```

### INTERPRETATION #1

There were 0 chloroplast ASVs and 0 mitochondrial ASVs detected in my dataset. This seems reasonable because my samples are rectal swab microbiome samples from human patients, which are expected to primarily contain bacterial DNA rather than plant-derived chloroplast sequences. Additionally, the study included contamination monitoring and filtering steps, which likely removed potential mitochondrial or chloroplast sequences if present.


# Evaluate and remove negative controls 
```{r}

# Check if there is negative control 

sample_names(noChlorosMitos_physeq)
```


### INTERPRETATION #2

No negative control samples were detected in my dataset. All sample names appear as SRR accession IDs, and none are labeled as controls.


# Positive Controls 

```{r}

# check if there is positive control 
sample_data(noChlorosMitos_physeq)
```


### INTERPRETATION #3

My dataset does not appear to include a mock community sample. The sample names in the downloaded dataset are all SRR accession IDs for human gut samples, and there is no sample labeled as a mock control. 

# Reverse Complements


```{r}
# Pull out the Taxa table 

tax_table <- 
  noChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV sequences and names 

asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix 

asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix names in rows and columns 

rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings 
asv_dna <- DNAStringSet(asv_seqs) 

# Take a look
head(asv_dna)

asv_rev_dna <- reverseComplement(asv_dna) 


for(i in 1:length(asv_seqs)){ 
  match_vec <- asv_dna[i] == asv_rev_dna 
  asv_rev_comp[,i] <- match_vec 
}

cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed")
```
### INTERPRETATION #4

There were 0 reverse complement ASVs detected in my dataset. This indicates that no ASV sequences were identified as reverse complements of each other, so no sequences need to be removed at this step. 


# Sequencing Depth 

```{r}
# Make a metadata dataframe for joining 

metadata_df <- 
  noChlorosMitos_physeq %>%
  sample_data() %>%
  data.frame() %>%
  tibble::rownames_to_column(var = "names") %>%
  dplyr::mutate(
    AttemptStatus = ifelse(
      grepl("non attempters", source_material_id),
      "Non-attempter",
      "Attempter"
    )
  )

# What is the library size/sequencing depth for each sample 

seqSums_df <- 
  noChlorosMitos_physeq %>%
  otu_table() %>%
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_df, by = "names")

# Rename second columns 

colnames(seqSums_df)[2] <- "TotalSeqs"

# Take a look 

dim(seqSums_df)
head(seqSums_df)

```

```{r}
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()
```


```{r}
# bar plot 
numSeq_bar_plot <- 
  seqSums_df %>%
  ggplot(aes(x = reorder(names, TotalSeqs), y = TotalSeqs,
             fill = AttemptStatus)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample") + 
  theme(legend.position = "none")

# histogram
numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x = TotalSeqs, fill = AttemptStatus)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme(legend.position = "bottom")

# Density plot 
numSeq_density_plot <- 
  seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = AttemptStatus)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth") + 
  theme(legend.position = "none")

numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A")
```



### INTERPRETATION #5

Sequencing depth varies across samples, with read counts ranging approximately from 12,000 to 47,000 reads. Most samples cluster between about 25,000 and 35,000 reads, indicating generally consistent sequencing depth across the dataset.

The distribution of sequencing depth appears similar between the attempter and non-attempter groups, as shown by the overlapping histograms and density curves. This suggests that sequencing depth is not strongly biased toward one group, reducing the likelihood that group differences are driven by uneven sequencing effort.

Overall, the sequencing depth across samples appears sufficient and relatively consistent, available for reliable downstream diversity analysis.


# Remove samples with few reads 

```{r}
# no longer needing to remove 
summary(seqSums_df$TotalSeqs)
#sum(seqSums_df$TotalSeqs < 100)

#seqSums_df %>%
  #dplyr::filter(TotalSeqs < 100) %>%
  #dplyr::select(names, TotalSeqs) %>%
  #dplyr::arrange(TotalSeqs)

#preprocessed_physeq <- 
 # prune_samples(
   # !(sample_names(noChlorosMitos_physeq) %in% c(
   #   "SRR35320078",
   #   "SRR35320071",
   #   "SRR35320081",
   #  "SRR35320073",
    #  "SRR35320124"
   # )),
   # noChlorosMitos_physeq
 # )

preprocessed_physeq <- noChlorosMitos_physeq
min(sample_sums(preprocessed_physeq))
nsamples(preprocessed_physeq)
```


```{r}

# Check for sequencing depth 

seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

# Write out the data 


```{r}
save(
  preprocessed_physeq,
  file = "data/preprocessed_phyloseq.RData"
)
```


# Final info for reproducibility 

```{r}

# Check Render Time 

end_time <- Sys.time()
end_time 
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
# Session Info 
devtools::session_info()
```

