---
title: "04_Biodiversity.Rmd"
author: "Morrow Zhang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

### Scientific Questions of Analysis 
Scientific Question: Do hospitalized patients with major depressive disorder (MDD) who have made a recent suicide attempt exhibit higher gut microbial alpha diversity compared to MDD patients without a history of suicide attempts? 

Alpha diversity Null hypothesis: There is no difference in gut microbial alpha diversity between hospitalized MDD patients with a recent suicide attempt and those without a history of suicide attempts.

Alternative hypothesis: Hospitalized MDD patients with a recent suicide attempt exhibit higher gut microbial alpha diversity compared to MDD patients without a history of suicide attempts.

# Set seed and Load Packages 

```{r setup}
start_time <- Sys.time()
set.seed(238428)
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

```

# Load in data and make metadata dataframe 
```{r}
# Load data
load("data/preprocessed_phyloseq.RData")
preprocessed_physeq

# Intuition check 
min(sample_sums(preprocessed_physeq))

# Create a metdata dataframe
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

# Take a look at the metadata dataframe
glimpse(metadata_df)
```


```{r}

# prepare iNEXT input data 
# Species in rows while samples in columns
iNEXT_input_df <- 
  preprocessed_physeq %>%
  otu_table() %>%
  data.frame()

# Take a look 

dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]
```

# Biodiversity Hill Numbers with iNEXT

```{r}
# Run iNEXT

#iNEXT_data <- iNEXT(
#  iNEXT_input_df,
#  q = c(0, 1, 2),
#  datatype = "abundance"
#)



#save(iNEXT_data, file = "data/iNEXT_data.RData")
```


```{r}

# Take a look 

load("data/iNEXT_data.RData")
str(iNEXT_data)
typeof(iNEXT_data)
```


# Rarefaction Curves

```{r}

# Colors Preparation
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

head(color_df)

colnames(color_df)[1] <- "names"

head(color_df)

# Make a helper dataframe for plotting with colors 

iNEXT_color_df <- 
  color_df %>%
  # Merge with metadata
  left_join(metadata_df, by = c("names" = "Run")) %>%
  # Add colors for plotting by your real grouping
  mutate(
    AttemptStatus = ifelse(
      grepl("non attempters", source_material_id),
      "Non-attempter",
      "Attempter"
    ),
    attempt_colors = ifelse(
      AttemptStatus == "Non-attempter",
      "blue",
      "red"
    )
  )

head(iNEXT_color_df)
```


```{r}
inext_plot <- ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") +
  scale_color_manual(values = iNEXT_color_df$attempt_colors) +
  scale_fill_manual(values = iNEXT_color_df$attempt_colors) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(y = "Effective Number of ASVs", x = "Number of Sequences")

inext_plot

# save my plot 
ggsave(
  filename = "../figures/04_Biodiversity/iNEXT_curve.png",
  plot = inext_plot,
  width = 12,
  height = 6,
  dpi = 300
)
```


# Manual Rarefaction Curves

```{r}
iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  left_join(metadata_df, by = c("names" = "Run")) %>%
  mutate(
    AttemptStatus = ifelse(
      grepl("non attempters", source_material_id),
      "Non-attempter",
      "Attempter"
    )
  )


dim(iNEXT_manual_df)

rarefaction_plot <- 
  iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolated)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot! 
  ggplot(aes(x = m, y = qD, color = AttemptStatus, group = names)) + 
  geom_line() + 
  facet_grid(Order.q ~ AttemptStatus, scales = "free") + 
  scale_color_manual(values = c(
    "Attempter" = "red",
    "Non-attempter" = "blue"
  )) + 
  theme_bw() + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") + 
  theme(legend.position = "bottom")

rarefaction_plot

ggsave(
  filename = "../figures/04_Biodiversity/manual_rarefaction_curves.png",
  plot = rarefaction_plot,
  width = 12,
  height = 8)
```


### Interpretation 1 
For the ggiNEXT and manual rarefaction curve plots, there should be enough sequences to analyze for q = 1 and q = 2, but not for q = 0 because many curves in q = 0 are still increasing and have not fully plateaued. Because of that, valid conclusion can be drawn on species richness when q = 1 and q = 2, because the curves stabilize clearly. However, it is probably not valid to draw conclusion when q = 0, because some curves are still increasing.


# Statistical Testing 

```{r}

# Tes if the data is normal 

distinct_metadata_df <- metadata_df
shapiro.test(distinct_metadata_df$AvgSpotLen)

obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed")
glimpse(obs_div_df)


```
```{r}

# Test if the data is normal

obs_whole_rich_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 0)

shapiro.test(obs_whole_rich_df$qD)
```
# p-value is less than 0.5 > my data is not normal 





```{r}
kw_attempt_rich <- 
  obs_whole_rich_df %>%
  kruskal_test(qD ~ AttemptStatus)

kw_attempt_rich
```

# Categorical Analysis 

```{r}
# Add the plot 
attempt_ggboxplot <- 
  ggboxplot(obs_whole_rich_df, x = "AttemptStatus", y = "qD", 
            color = "AttemptStatus", fill = "AttemptStatus", alpha = 0.3,
            outlier.shape = NA) + 
  # Add points
  geom_jitter(aes(color = AttemptStatus)) + 
  scale_color_manual(values = c("Attempter" = "red",
                                "Non-attempter" = "blue")) + 
  scale_fill_manual(values = c("Attempter" = "red",
                               "Non-attempter" = "blue")) + 
  labs(y = "# of Total ASVs (Richness)") + 
  # Add the KW results 
  labs(subtitle = get_test_label(kw_attempt_rich, detailed = TRUE)) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))

# Show the plot

attempt_ggboxplot

# save the alpha richness boxplot
ggsave(
  filename = "../figures/04_Biodiversity/alpha_richness_boxplot.png",
  plot = attempt_ggboxplot,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r}
# Plot boxplots 

alpha_all_q_plot <- 
  obs_div_df %>%
  ggplot(aes(x = AttemptStatus, y = qD, fill = AttemptStatus, 
             color = AttemptStatus)) + 
  facet_wrap(.~Order.q, scales = "free_y") + 
  geom_jitter() + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = c("Attempter" = "red",
                                "Non-attempter" = "blue")) + 
  scale_fill_manual(values = c("Attempter" = "red",
                               "Non-attempter" = "blue")) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))

alpha_all_q_plot
ggsave(
  filename = "../figures/04_Biodiversity/alpha_diversity_all_q.png",
  plot = alpha_all_q_plot,
  width = 12,
  height = 6,
  dpi = 300
)
```
```{r q = 0}
# I use Wilcox instead of ANOVA, because my data is not normal.

obs_rich_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 0)


wilcox_attempt_rich <- 
  wilcox_test(qD ~ AttemptStatus, data = obs_rich_df, alternative = "greater")


wilcox_attempt_rich
```
```{r q = 1}
obs_shannon_df <-
  obs_div_df %>%
  dplyr::filter(Order.q == 1)

wilcox_attempt_shannon <-
  wilcox_test(qD ~ AttemptStatus,
               data = obs_shannon_df, alternative = "greater")

wilcox_attempt_shannon
```

```{r q = 2}
obs_simpson_df <-
  obs_div_df %>%
  dplyr::filter(Order.q == 2)

wilcox_attempt_simpson <-
  wilcox_test(qD ~ AttemptStatus,
               data = obs_simpson_df, alternative = "greater")

wilcox_attempt_simpson
```

```{r}
obs_simps_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 2)
obs_simps_df
```

```{r}

#Rich 
rich_ggboxplot <- 
  ggboxplot(obs_rich_df, x = "AttemptStatus", y = "qD", 
            color = "AttemptStatus", fill = "AttemptStatus", alpha = 0.3,
            outlier.shape = NA) + 
 
  geom_jitter(aes(color = AttemptStatus)) + 
  scale_color_manual(values = c("Attempter" = "red",
                                "Non-attempter" = "blue")) + 
  scale_fill_manual(values = c("Attempter" = "red",
                               "Non-attempter" = "blue")) + 
  labs(y = "# of Total ASVs (Richness)") + 
  # Add Wilcoxon
  stat_compare_means(method = "wilcox.test", label.y = min(obs_rich_df$qD) - 5) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))


rich_ggboxplot

ggsave(
  filename = "../figures/04_Biodiversity/richness_wilcoxon_boxplot.png",
  plot = rich_ggboxplot,
  width = 8,
  height = 6,
  dpi = 300
)
```


```{r}
# Simpson by AttemptStatus 
simps_ggboxplot <- 
  ggboxplot(obs_simps_df, x = "AttemptStatus", y = "qD", 
            color = "AttemptStatus", fill = "AttemptStatus", alpha = 0.3,
            outlier.shape = NA) + 


  geom_jitter(aes(color = AttemptStatus)) + 

  scale_color_manual(values = c("Attempter" = "red",
                                "Non-attempter" = "blue")) + 

  scale_fill_manual(values = c("Attempter" = "red",
                               "Non-attempter" = "blue")) + 

  labs(y = "# of Dominant ASVs (Simpson)") +

  # Add Wilcoxon
  stat_compare_means(method = "wilcox.test",
                     label.y = min(obs_simps_df$qD) - 2) + 

  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1,
                                   vjust = 1))

simps_ggboxplot

ggsave(
  filename = "../figures/04_Biodiversity/simpson_wilcoxon_boxplot.png",
  plot = simps_ggboxplot,
  width = 8,
  height = 6,
  dpi = 300
)
```


```{r}
rich_ggboxplot + simps_ggboxplot
```


### Interpretation 2
Wilcoxon rank-sum tests using a one-sided alternative hypothesis showed no significant differences in alpha diversity between suicide attempters and non-attempters across all diversity levels. When q = 0, p value is 0.11. When q = 1, p value is 0.237. When q = 2, p value is 0.443. Overall, these results fail to reject the null hypothesis and indicate that hospitalized MDD patients with recent suicide attempts do not exhibit significantly higher gut microbial alpha diversity compared to those without a history of suicide attempts.


### Interpretation 3
I statistically tested my scientific question using Wilcoxon rank-sum tests to compare alpha diversity between suicide attempters and non-attempters. I used this test because my variable of interest, Attempt Status, is categorical, and my alpha diversity data were not normally distributed, making a non-parametric test more appropriate than ANOVA. I conducted separate one-sided Wilcoxon tests for richness (q = 0), Shannon diversity (q = 1), and Simpson diversity (q = 2), using an alternative hypothesis that alpha diversity would be greater in attempters compared to non-attempters. This allowed me to statistically evaluate whether gut microbial diversity differed between groups across all ASVs, common ASVs, and dominant ASVs.

# Info for Reproducibility 

```{r}
end_time <- Sys.time()
end_time 
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
devtools::session_info()
```

