---
title: "05_Ordination.Rmd"
author: "Morrow Zhang"
date: "`r Sys.Date()`"
output: html_document
---
# Goal

Load in phyloseq data with rooted tree and colors and functions.
Evaluate sequencing depth and remove samples, if needed.
Normalize the read counts between samples.
Calculate community dissimilarities. Numbers between 0 and 1. If 0, completely similar versus if they are 1, then they’re completely dissimilar.
Sorensen: Shared Species as a binary value: Abundance-unweighted
Bray-Curtis: Shared Abundant species: Abundance-weighted
(Abundance-)Weighted UNIFRAC: Consider Abundant Species and where they fall on the tree
Visualize the community data with two unconstrained Ordinations:
PCoA: Linear Method. Eigenvalue = how much variation is explained by each axis. Choose to view axis 1, 2, 3, etc. and plot them together.
NMDS: Non-linear. Smush multiple Dimensions into 2 or 3 axes. Need to report Stress value (ideally <0.15).
Run statistics with functions from the vegan R package:
PERMANOVA with adonis2().
betadispR with betadisper() and permutest().

# Scientific Question 
### Interpretation 1 
Scientific Question: Do hospitalized patients with MDD who have made a recent suicide attempt have distinct microbial community composition compared to MDD patients without a history of suicide attempts?

Null Hypothesis: Gut microbial community composition does not differ between hospitalized MDD patients with a recent suicide attempt and those without a history of suicide attempts.

Alternative Hypothesis: Hospitalized MDD patients with a recent suicide attempt have significantly different microbial community composition compared to MDD patients without a history of suicide attempts.

These hypotheses relate to both abundance-unweighted and abundance-weighted-beta diversity metrics. Abundance-unweighted metrics test whether different microbial taxa are present between attempters and non-attempters, while abundance-weighted metrics test whether dominant taxa differ in relative abundance between groups.


# Set up

```{r setup, include=FALSE}
start_time <- Sys.time()

# set seed
set.seed(238428)

# load package 
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, 
               install = FALSE)

source("../functions.R")
```

# load in data 

```{r}

# load phyloseq object
load("../data/preprocessed_phyloseq.RData")
preprocessed_physeq
min(sample_sums(preprocessed_physeq))
```
# Remove outlier if any 

```{r}
# find if there is any outlier 

sort(sample_sums(preprocessed_physeq))
```


```{r}

# create metadata frame 
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()
```


# Normalize the Read Depth 

## Explore the raw read counts

```{r}
raw_TotalSeqs_df <- 
  preprocessed_physeq %>%
  # Calculate the total number of sequences/reads
  sample_sums() %>%
  data.frame()

# Take a look 
head(raw_TotalSeqs_df)
```


```{r}
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
raw_TotalSeqsASVs_df <- 
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(preprocessed_physeq) > 1))

# View(raw_TotalSeqsASVs_df)

#  histogram of seq depth 
rawSeq_histogram <-
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram(bins = 50) +
  labs(title = "Raw Seq Depth Histogram") +
  theme_bw()

# Seq Depth versus num_ASVs
rawSeq_vs_numASV_plot <-
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Raw Seq Depth vs # ASVs") +
  theme_bw()

# test significance? 
summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))
```
```{r}

# put plots together 

rawSeq_histogram + rawSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```


# Scale the Read Counts

```{r}
min(sample_sums(preprocessed_physeq))

# Scale the reads 

scaled_physeq <- 
  preprocessed_physeq %>%
  scale_reads(round = "matround")

# Look at it 
scaled_physeq
```


```{r}
# Look at it more!
#View(data.frame(otu_table(preprocessed_physeq)))
#View(data.frame(otu_table(scaled_physeq)))

# Confirm seq read depth of scaled_physeq 
scaled_TotalSeqs_df <- 
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
scaled_TotalSeqsASVs_df <- 
  scaled_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(scaled_physeq) > 1))

# Plot it! 
scaledSeq_histogram <- 
  scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  labs(title = "Scaled Seq Depth Histogram") + 
  #scale_x_continuous(limits = c(0, 7500)) + 
  theme_bw()

# Scaling always gives us a range! 
# MInimum sequences
min(scaled_TotalSeqs_df$TotalSeqs) 
```

```{r}
# Total Range 
range(scaled_TotalSeqs_df)

# How many is that range? 
range_seqs <- (max(scaled_TotalSeqs_df$TotalSeqs) - min(scaled_TotalSeqs_df$TotalSeqs))
range_seqs

# And in terms of percent? 
range_seqs/max(scaled_TotalSeqs_df$TotalSeqs)
```


```{r}
#View(scaled_TotalSeqs_df)

# Set the axis min and max for the next plot 
max_y <- max(scaled_TotalSeqs_df$TotalSeqs) + 10
min_y <- min(scaled_TotalSeqs_df$TotalSeqs) - 10

# Now, draw the plot 
scaledSeq_vs_numASV_plot <- 
  scaled_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(min_y, max_y)) +
  #geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(title = "Scaled Seq Depth vs # ASVs")

# Put the two plots together.
scaledSeq_histogram + scaledSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```

# Calculating dissimarity 

```{r}
# Sorensen Dissimiliarty
scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

# What does it look like? 
class(scaled_sorensen_dist)

str(scaled_sorensen_dist)

#head(as.matrix(scaled_sorensen_dist))

# Bray-Curtis Dissimiliarty
scaled_bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)

# Abundance-Unweighted UniFrac
#scaled_uUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "unifrac")

# Abundance-Weighted UniFrac
#scaled_wUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "wunifrac")

```

# Visualize Community Dissimilarity with Ordination

```{r}
# First, calculate PCoA with Soresen
scaled_soren_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE)

# Take a quick look
str(scaled_soren_pcoa)

```
```{r}
sample_data(preprocessed_physeq)$AttemptStatus <-
  ifelse(
    grepl("non attempter|non attempters", sample_data(preprocessed_physeq)$source_material_id, ignore.case = TRUE),
    "Non-attempter",
    ifelse(
      grepl("attempter|attempters", sample_data(preprocessed_physeq)$source_material_id, ignore.case = TRUE),
      "Attempter",
      NA
    )
  )
```


```{r}
table(sample_data(preprocessed_physeq)$AttemptStatus, useNA = "ifany")
```


```{r}
attempt_colors <- c(
  "Attempter" = "#D55E00",
  "Non-attempter" = "#0072B2"
)

soren_pcoa_plot <-
  plot_ordination(
    physeq = preprocessed_physeq,
    ordination = soren_pcoa,
    color = "AttemptStatus",
    title = "Sorensen PCoA"
  ) +
  geom_point(size = 5, alpha = 0.7) +
  scale_color_manual(values = attempt_colors) +
  theme_bw() +
  theme(legend.position = "right")

bray_pcoa_plot <-
  plot_ordination(
    physeq = preprocessed_physeq,
    ordination = bray_pcoa,
    color = "AttemptStatus",
    title = "Bray-Curtis PCoA"
  ) +
  geom_point(size = 5, alpha = 0.7) +
  scale_color_manual(values = attempt_colors) +
  theme_bw() +
  theme(legend.position = "right")

soren_pcoa_plot + bray_pcoa_plot +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

```{r}
sample_data(scaled_physeq)$AttemptStatus <-
  sample_data(preprocessed_physeq)$AttemptStatus
table(sample_data(scaled_physeq)$AttemptStatus, useNA = "ifany")
```


# NMDS: Non-Metric Multidimensional Scaling

```{r}
## SORENSEN 


attempt_colors <- c(
  "Attempter" = "#D55E00",
  "Non-attempter" = "#0072B2"
)

scaled_soren_nmds <-
  ordinate(
    physeq = scaled_physeq,
    method = "NMDS",
    distance = "bray",
    binary = FALSE
  )

soren_nmds_plot <-
  plot_ordination(
    physeq = scaled_physeq,
    ordination = scaled_soren_nmds,
    color = "AttemptStatus",
    title = "Sorensen NMDS"
  ) +
  scale_color_manual(values = attempt_colors) +
  geom_point(size = 5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right")
### Bray-Curtis
# Second, calculate NMDS with Bray-Curtis
bray_nmds <-
  ordinate(
    physeq = preprocessed_physeq,
    method = "NMDS",
    distance = "bray",
    binary = FALSE
  )

# Plot it: Bray-Curtis NMDS
bray_nmds_plot <-
  plot_ordination(
    physeq = preprocessed_physeq,
    ordination = bray_nmds,
    color = "AttemptStatus",
    title = "Bray-Curtis NMDS"
  ) +
  scale_color_manual(values = attempt_colors) +
  geom_point(size = 5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right")

# Show the plots
soren_nmds_plot + bray_nmds_plot +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

# All ordinations together 

```{r}
soren_pcoa_plot + bray_pcoa_plot + 
soren_nmds_plot + bray_nmds_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```
```{r}
## 1. Run with by = terms for R² values, sensitive to order of variables! 
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
sorensen_attempt_adonis <- adonis2(
  scaled_sorensen_dist ~ AttemptStatus,
  data = data.frame(sample_data(scaled_physeq)),
  by = "terms"
)

sorensen_attempt_adonis
```
```{r}
sorensen_attempt_adonis_terms2 <- 
  adonis2(
    scaled_sorensen_dist ~ AttemptStatus,
    data = data.frame(sample_data(scaled_physeq)),
    by = "terms"
  )

sorensen_attempt_adonis_terms2
```
```{r}
# Bray-Curtis
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
bray_attempt_adonis_terms1 <- 
  adonis2(
    scaled_bray_dist ~ AttemptStatus,
    data = data.frame(sample_data(scaled_physeq)),
    by = "terms"
  )

bray_attempt_adonis_terms1
```
```{r}
bray_attempt_adonis_terms2 <- 
  adonis2(
    scaled_bray_dist ~ AttemptStatus,
    data = data.frame(sample_data(scaled_physeq)),
    by = "terms"
  )

bray_attempt_adonis_terms2
```
```{r}
# Homogeneity of Dispersion test with beta disper
# Sorensen Betadisper - AttemptStatus (Categorical)

dispr_sorensen_attempt <- 
  betadisper(
    scaled_sorensen_dist,
    sample_data(scaled_physeq)$AttemptStatus
  )

# Permutation test
permutest(dispr_sorensen_attempt)
```

```{r}
# Sorensen Betadisper - AttemptStatus  

dispr_sorensen_attempt2 <- 
  betadisper(
    scaled_sorensen_dist,
    sample_data(scaled_physeq)$AttemptStatus
  )

permutest(dispr_sorensen_attempt2)
```
```{r}
dispr_sorensen_attempt3 <- 
  betadisper(
    scaled_sorensen_dist,
    sample_data(scaled_physeq)$AttemptStatus
  )

permutest(dispr_sorensen_attempt3)
```
```{r}
# Bray-Curtis Betadisper - AttemptStatus (Categorical)

dispr_bray_attempt <- 
  betadisper(
    scaled_bray_dist,
    sample_data(scaled_physeq)$AttemptStatus
  )

permutest(dispr_bray_attempt)
```
```{r}
end_time <- Sys.time()
end_time 

elapsed_time <- round((end_time - start_time), 3)
elapsed_time

# Ensure reproducibility with package version information
devtools::session_info()
```

