---
title: "05_Ordination.Rmd"
author: "Morrow Zhang"
date: "`r Sys.Date()`"
output: html_document
---
# Goal

Load in phyloseq data with rooted tree and colors and functions.
Evaluate sequencing depth and remove samples, if needed.
Normalize the read counts between samples.
Calculate community dissimilarities. Numbers between 0 and 1. If 0, completely similar versus if they are 1, then they’re completely dissimilar.
Sorensen: Shared Species as a binary value: Abundance-unweighted
Bray-Curtis: Shared Abundant species: Abundance-weighted
(Abundance-)Weighted UNIFRAC: Consider Abundant Species and where they fall on the tree
Visualize the community data with two unconstrained Ordinations:
PCoA: Linear Method. Eigenvalue = how much variation is explained by each axis. Choose to view axis 1, 2, 3, etc. and plot them together.
NMDS: Non-linear. Smush multiple Dimensions into 2 or 3 axes. Need to report Stress value (ideally <0.15).
Run statistics with functions from the vegan R package:
PERMANOVA with adonis2().
betadispR with betadisper() and permutest().

# Scientific Question 
### Interpretation 1 
Scientific Question: Do hospitalized patients with MDD who have made a recent suicide attempt have distinct microbial community composition compared to MDD patients without a history of suicide attempts?

Null Hypothesis: Gut microbial community composition does not differ between hospitalized MDD patients with a recent suicide attempt and those without a history of suicide attempts.

Alternative Hypothesis: Hospitalized MDD patients with a recent suicide attempt have significantly different microbial community composition compared to MDD patients without a history of suicide attempts.

These hypotheses relate to both abundance-unweighted and abundance-weighted-beta diversity metrics. Abundance-unweighted metrics test whether different microbial taxa are present between attempters and non-attempters, while abundance-weighted metrics test whether dominant taxa differ in relative abundance between groups.


```{r setup, include=FALSE}
start_time <- Sys.time()

set.seed(238428)

pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, 
               install = FALSE)


```

```{r}
# load phyloseq object
load("../data/preprocessed_phyloseq.RData")
preprocessed_physeq
min(sample_sums(preprocessed_physeq))
```
```{r}
# find if there is any outlier 

sort(sample_sums(preprocessed_physeq))
```
```{r}
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()
```

```{r}
raw_TotalSeqs_df <- 
  preprocessed_physeq %>%
  # Calculate the total number of sequences/reads
  sample_sums() %>%
  data.frame()

# Take a look 
head(raw_TotalSeqs_df)
```


```{r}
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
raw_TotalSeqsASVs_df <- 
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(preprocessed_physeq) > 1))

# View(raw_TotalSeqsASVs_df)

#  histogram of seq depth 
rawSeq_histogram <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  scale_x_continuous(limits = c(0, 7500)) + 
  labs(title = "Raw Seq Depth Histogram") + 
  theme_bw()

# Seq Depth versus num_ASVs
rawSeq_vs_numASV_plot <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 7500)) +
  geom_smooth(method = "lm") + 
  labs(title = "Raw Seq Depth vs # ASVs") + 
  theme_bw()

# test significance? 
summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))
```
```{r}
rawSeq_histogram + rawSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```

```{r}
min(sample_sums(preprocessed_physeq))
```

```{r}

```

